---
title: "Applied Statistics. HW1"
author: "Rutkovskaya Ekaterina"
date: "2024-03-29"
output: html_document
---
<style type="text/css">
  body{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ordination and community analysis


```{r, message=FALSE, warning=FALSE}
library(readxl)
library(tidyr)
library(dplyr)
library(broom)
library(ggmap)
library(ggvegan)
library(vegan)
library(ggplot2)

data("BCI")
data('BCI.env')
```

### EDA

Counting how many species in each of 50 tree plots 
```{r, message=FALSE, warning=FALSE}
BCI$Diversity <- rowSums(BCI > 0)

BCI %>% ggplot(aes(x=rownames(BCI), y=BCI$Diversity, fill=factor(BCI.env$Habitat))) +
  xlab("Tree plots") + ylab("Number of species") + labs(fill = "Habitat") +
  geom_col() + coord_flip() + scale_fill_brewer(palette = "Pastel1")
```


Finding the lead tree species in each plot
```{r, message=FALSE, warning=FALSE}
max_col_index <- max.col(BCI[, c(0:225)])
sp_name <- colnames(BCI)[max_col_index]
BCI$Indiv_number <- apply(BCI[, -ncol(BCI)], 1, max)
BCI$Dominant_sp <-sp_name

BCI %>% ggplot(aes(x = factor(row.names(BCI)), fill = factor(Dominant_sp), y = Indiv_number)) +
  xlab("Tree plots") + ylab("Number of individuals") + labs(fill = "Species") +
  geom_dotplot(binaxis = "y", stackdir = "center", position = "dodge", binwidth = 5) + scale_fill_brewer(palette = "Set2")
```

Data preparation
```{r, message=FALSE, warning=FALSE}
BCI[nrow(BCI) + 1,] = apply(BCI, 2, max)
low_number<- as.data.frame(lapply(BCI[, BCI[51, ] < 20], as.integer))
low_number$Total <-rowSums(low_number)
BCI<-as.data.frame(lapply(BCI[, BCI[51, c(1:225)] > 20], as.integer))
BCI$Other <- low_number$Total
BCI<-BCI[c(1:50), ]
```

nMDS
```{r, message=FALSE, warning=FALSE}
ordination <- metaMDS(BCI, dist = "bray", autotransform = FALSE)
autoplot(ordination)
```

```{r, message=FALSE, warning=FALSE}
fort <-fortify(ordination)
ggplot()+
  geom_point(data=subset(fort, score =='sites'),
             mapping = aes(x=NMDS1, y=NMDS2),
             colour='purple',
             alpha=0.5)+
  geom_segment(data=subset(fort, score='species'),
               mapping=aes(x=0, y=0, xend=NMDS1, yend=NMDS2),
               arrow=arrow(length = unit(0.015, 'npc'),
                           type='closed'),
               colour='darkgray',
               size=0,
               alpha=0)+
  geom_text(data=subset(fort, score='species'),
            mapping=aes(label=label, x=NMDS1*1.1, y=NMDS2*1.1),
            alpha=0)+
  geom_abline(intercept = 0, slope = 0, linetype='dashed', size=0.8, colour='gray')+
  geom_vline(aes(xintercept=0), linetype='dashed', size=0.8, colour='gray')+
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank(),
        axis.line = element_line(colour = 'black'))
```

In summary we can see that non of the environmental factors are equally distributed, so we can use Habitat as a factor to explore. 
```{r, message=FALSE, warning=FALSE}
summary(BCI.env)
```

We can see the clear separation between OldSlope and OldHigh populations:
```{r, message=FALSE, warning=FALSE}
adonis2(BCI~Habitat, data=BCI.env)
ggplot()+
  geom_point(data=subset(fort, score =='sites'),
             mapping = aes(x=NMDS1, y=NMDS2, colour=BCI.env$Habitat),
             alpha=0.5)+ 
  geom_segment(data=subset(fort, score='species'),
               mapping=aes(x=0, y=0, xend=NMDS1, yend=NMDS2),
               arrow=arrow(length = unit(0.015, 'npc'),
                           type='closed'),
               colour='darkgray',
               size=0,
               alpha=0)+
  geom_text(data=subset(fort, score='species'),
            mapping=aes(label=label, x=NMDS1*1.1, y=NMDS2*1.1),
            alpha=0)+
  geom_abline(intercept = 0, slope = 0, linetype='dashed', size=0.8, colour='gray')+
  geom_vline(aes(xintercept=0), linetype='dashed', size=0.8, colour='gray')+
  theme(panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.background = element_blank(),
      axis.line = element_line(colour = 'black'),
      legend.position = c(0.8, 0.2))+
  scale_color_discrete("Habitat")
```

Interpretation of ordination
```{r, message=FALSE, warning=FALSE}
fit <- envfit(ordination, BCI.env, perm = 999)
fit$factors
fit$vectors

pal_sh <- c(1, 2)
ordiplot(ordination, type = "n")
points(ordination, col = BCI.env$Habitat, pch = pal_sh[BCI.env$Stream])
plot(fit)

```

We can see some correlation between Environmental Heterogeneity and UTM coordinates of the trees, but there is almost negative correlation between E-W and N-S distribution of species. The rest factors are not fitted well. 


All three factors have almost linear relationship to the data, but North-South distribution tends more to non-linear way.
```{r, message=FALSE, warning=FALSE}
os_EH <- ordisurf(ordination, BCI.env$EnvHet, method = "REML", main="Envitonmental Heterogenity", xlab="NMDS1", ylab="NMDS2", bubble=5)
summary(os_EH)


os_coord_EW<-ordisurf(ordination, BCI.env$UTM.EW, method = "REML",  main="East-West coordinates", xlab="NMDS1", ylab="NMDS2", bubble=5)
summary(os_coord_EW)


os_coord_NS<-ordisurf(ordination, BCI.env$UTM.NS, method = "REML",  main="North-South coordinates", xlab="NMDS1", ylab="NMDS2", bubble=5)
summary(os_coord_NS)

```
