---
title: "Applied Statistics. HW3"
author: "Rutkovskaya Ekaterina"
date: "2024-05-14"
output: html_document
---

```{r, message=FALSE, warning=FALSE}
library(readxl)
library(vegan)
library(ggplot2)
library(plotly)
library(impute)
library(factoextra)
library(psych)
library(ggforce)
library(rstatix)
library(ggsignif)
library(gridExtra)
```

## Data uploading processing
```{r, message=FALSE, warning=FALSE}
df <- read_excel("C:/Users/rutko/Downloads/Sleepy lizard.xlsx")
df <- df %>% 
  select(Treatment, Habitat, Connectivity, Tot_WBC, Het_ABS, Lym_ABS, `H:L Ratio`, Mon_ABS, OthG_ABS, LBSI)

df$Treatment <-  as.factor(df$Treatment)
df$Habitat <- as.factor(df$Habitat)
df$Connectivity <-  as.factor(df$Connectivity)
```

## EDA
```{r, message=FALSE, warning=FALSE}
colSums(is.na(df))
```

```{r, message=FALSE, warning=FALSE}
blood_parameters <-  df[, 4:10]
pairs(blood_parameters, pch = 19, cex=0.5)
```

```{r, message=FALSE, warning=FALSE}
plot_with_significance <- function(data, column_name, y_label) {
  ggplot(data, aes(x = Treatment, y = !!sym(column_name), fill = Treatment)) +
    geom_boxplot() +
    geom_signif(comparisons = list(c("1", "2")), 
                map_signif_level = TRUE, 
                test = "t.test") +
    labs(y = y_label, title = paste(y_label)) +
    theme_minimal()
}

columns_to_plot <- colnames(blood_parameters)
labels <- colnames(blood_parameters)


plots <- lapply(1:length(columns_to_plot), function(i) {
  plot_with_significance(df, columns_to_plot[i], labels[i])
})

grid.arrange(grobs = plots, ncol = 2)
```

Let's check data distribution
```{r, message=FALSE, warning=FALSE}
boxplot(blood_parameters)
```

Looks like we need some normalization
```{r, message=FALSE, warning=FALSE}
blood_parameters <- as.data.frame(scale(log(blood_parameters+1)))
normalized_data <- cbind(df[, 1:3], blood_parameters)
boxplot(blood_parameters)
```

That's better

```{r, message=FALSE, warning=FALSE}
pca_df <- prcomp(blood_parameters)$x %>% 
  as.data.frame() %>%
  select(PC1, PC2) %>% 
  mutate(Treatment = normalized_data$Treatment,
         Habitat = normalized_data$Habitat)

ggplot(pca_df, aes(x=PC1, y=PC2, shape=Treatment, col=Habitat))+
  geom_point(size=3)+
  geom_mark_ellipse(aes(col = Treatment, fill=Treatment))
```

As expected, we have two main classes. Let's find out how they differ.

## Checking dispersion
```{r, message=FALSE, warning=FALSE}
distances <- vegdist(blood_parameters, method  = "euclidean")
PCO_treat<- betadisper(distances, normalized_data$Treatment)
plot(PCO_treat)
anova(PCO_treat)
```

```{r, message=FALSE, warning=FALSE}
distances <- vegdist(blood_parameters, method  = "euclidean")
PCO_hab<- betadisper(distances, normalized_data$Habitat)
plot(PCO_hab)
anova(PCO_hab)
```

```{r, message=FALSE, warning=FALSE}
distances <- vegdist(blood_parameters, method  = "euclidean")
PCO_con<- betadisper(distances, normalized_data$Connectivity)
plot(PCO_con)
anova(PCO_con)
```

Variances in groups are significantly different, so alongside perMANOVA I'll use redundancy analysis

## Blood composition of different landscapes 
```{r, message=FALSE, warning=FALSE}
permanova_treat <- adonis2(blood_parameters ~ normalized_data$Treatment, data = normalized_data, permutations = 999, method = "euclidean")
print(permanova_treat)
```

```{r, message=FALSE, warning=FALSE}
pcoa_treat <- capscale(blood_parameters ~ normalized_data$Treatment, distance = "euclidean")
anova(pcoa_treat, by = "term", permutations = 999)
```

## Blood composition of different habitats

```{r, message=FALSE, warning=FALSE}
modified_blood <- blood_parameters%>% filter(normalized_data$Treatment == 2)
habitats <- normalized_data[normalized_data$Treatment == 2,]$Habitat
permanova_hab<-adonis2(modified_blood ~ habitats, method = "euclidean")
print(permanova_hab)
```

```{r, message=FALSE, warning=FALSE}
pcoa_hab <- capscale(modified_blood ~ habitats, distance = "euclidean")
anova(pcoa_hab, by = "term", permutations = 999)
```

## Blood composition of different connectivity
```{r, message=FALSE, warning=FALSE}
connectivity <- subset(normalized_data, Treatment == 2)$Connectivity
permanova_con<-adonis2(modified_blood ~ connectivity, method = "euclidean")
print(permanova_con)
```

```{r, message=FALSE, warning=FALSE}
pcoa_con <- capscale(modified_blood ~ connectivity, distance = "euclidean")
anova(pcoa_con, by = "term", permutations = 999)
```

## Task 2
```{r, message=FALSE, warning=FALSE}
permanova_multi<-adonis2(modified_blood~ connectivity + habitats, method = "euclidean")
print(permanova_multi)
```

```{r, message=FALSE, warning=FALSE}
pcoa_multi <- capscale(modified_blood ~ connectivity + habitats, distance = "euclidean")
anova(pcoa_multi, by = "term", permutations = 999)
```
Two tests (perMANOVA and db-RDA) showed the same results: treatment is only significant factor that influences blood composition of lizards. Habitat and connectivity nor them both have no significant effect on lizards.