---
title: "HW5"
author: "Rutkovskaya Ekaterina"
date: "2024-06-02"
output: html_document
---
## Data uploading
```{r, message=FALSE, warning=FALSE}
library(ade4)
library(readxl)
library(vegan)
library(ggplot2)
library(plotly)
library(impute)
library(factoextra)
library(psych)
library(ggforce)
library(rstatix)
data(doubs)
fish<- doubs$fish
env <- doubs$env
coord<-doubs$xy
```

## EDA

After some data exploring I noticed that not all sites have fish, so I decided to remove this sites
```{r, message=FALSE, warning=FALSE}
keep <- apply(fish, 1, sum) > 0
fish <- fish[keep, ]
env <- env[keep, ]
coord <- coord[keep, ]
```

## CCA

Now let's perform CCA for all environmental factors
```{r, message=FALSE, warning=FALSE}
cca_env <- cca(fish ~ ., env)
vif.cca(cca_env)
```
Variance inflation factor for distance (dfs) is too high, so let's remove this factor 

```{r, message=FALSE, warning=FALSE}
no_dfs<-env[,2:11]
cca_env <- cca(fish ~ ., no_dfs)
vif.cca(cca_env)
```
Now the factors are okay and I can continue analysis

```{r, message=FALSE, warning=FALSE}
summary(cca_env)
```

About 70% of the total variance is explained by the environmental variables, which is pretty good, and now I can plot the canonical axes

```{r, message=FALSE, warning=FALSE}
plot(cca_env, scaling = 1, 
     main = 'Biplot with Eucledian distance')
```

```{r, message=FALSE, warning=FALSE}
plot(cca_env, scaling = 2, 
     display = c("species", "cn"), 
     main = "Biplot with species-focused scaling")
```

On biplots I see that the most species are influenced by stream flow, hardness of water and nitrogen levels. Also there are two separate groups: Thymallus thymallus (Thth), Cottus gobio (Cogo) and Telestes soufia agassizi (Teso) are more present in high oxygen environment; Phoxinus phoxinus (Phph), Nemacheilus barbatulus (Nebo) and Salmo trutta fario (Satr) prefer the high slope and altitude places.   

```{r, message=FALSE, warning=FALSE}
anova(cca_env)
```

The p-value for the model is very high, so all the environmental variables collectively explain a significant amount of the variation in the fish species data.

Now let's find out which of the factors is the most influential

```{r, message=FALSE, warning=FALSE}
anova(cca_env, by="term")
```

Looks like the altitude and dissolved oxygen are two most significant predictors of fish community structure. Flow, pH, total hardness of water and phosphates concentration influence the community too. 

